---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# bathytools

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/limnotrack/bathytools/graph/badge.svg)](https://app.codecov.io/gh/limnotrack/bathytools)
[![R-CMD-check](https://github.com/limnotrack/bathytools/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/limnotrack/bathytools/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of bathytools is to facilitate the generation of bathymetric data for lakes and reservoirs. The package provides functions to generate depth points from bathymetric data, and to rasterise these points into a bathymetric raster. The package is designed to work with data from the AEME project, but can be used with any bathymetric data.

## Installation

You can install the development version of bathytools from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("limnotrack/bathytools")
```

## Example

This is a basic example which shows you how to generate a bathymetric raster from a shoreline and depth points:

```{r example}
library(bathytools)
## basic example code
shoreline <- readRDS(system.file("extdata/rotoma_shoreline.rds",
                                 package = "bathytools"))
point_data <- readRDS(system.file("extdata/depth_points.rds",
                                  package = "bathytools"))
bathy <- rasterise_bathy(shoreline = shoreline, point_data = point_data,
                         crs = 2193)
```
